A Plasma Torus Around a Young Low-Mass Star

### 2025-05-29 paper/ms.tex
update

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.925\textwidth]{f2.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Evolution of TIC\,141146667 during the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps were caused by light from the Earth (BTJD 3356-3358.5) and
  Moon (BTJD 3359.5-3360.5).  Raw two minute data are in gray; black
  time-averages to ten minute sampling.  {\bf b-c,} Folded TESS light
  curve before and after spectroscopy.  Black now phase-averages to
  100 points per 3.93\ hour cycle.  During BTJD 3352-3356, a state
  switch occurred near BTJD 3353, and the dip at $\phi$$\approx$0.8
  \added{became less pronounced}.  While the \added{dip spanning
  $\phi$=0.6-1.1} was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{f3.pdf}
  \vspace{-0.3cm}
  \captionsetup{labelformat=moviefmt,labelsep=colon}
	\caption{\textbf{Emission from plasma clumps orbiting
TIC\,141146667.}
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{{\bf
  available here}},
  shows the spectral evolution over five hours.
  {\bf a,} TESS light curve from 5 February 2024 to 26
  February 2024 folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra acquired on 17 February 2024.
  The continuum is set to unity, and the darkest color is at twice the
  continuum to accentuate emission outside the line core ($|v/v_{\rm
  eq}|>1$, for the equatorial velocity $v_{\rm eq}$=130\,\kms).  While
  emission in the line core comes from the star's chromosphere, the
  sinusoidal emission features are most readily described by a warped
  plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the online movie.
  The dotted line shows the time-averaged spectrum, $f_{\langle t
  \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting $f_{\langle t
  \rangle}$.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch mirrors the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.  \added{A separate version with overplotted best-fit sinusoids is also
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel_sinusoids.mp4}{{\bf
  available}}.}}
  \label{fig:spec}
\end{figure*}

@@

In a follow-up study, \citet{Townsend2008} made synthetic light
curves, assuming that the optical depth scaled linearly with plasma
density.  They found that W-shaped eclipses, similar to those seen for
TIC\,141146667, can occur when the spin and magnetic axes are
moderately (15-45$^\circ$) misaligned.  The movies associated with
their work are available
online\footnote{\url{http://user.astro.wisc.edu/\~townsend/static.php?ref=rrm-movies\#Download\_Bundles}
last accessed 23 April 2025.}; for a $\omega/\omega_{\rm
c}$$\approx$0.5 edge-on system like TIC\,141146667, the photometric
eclipses have the correct shape, and the H$\alpha$ emission similarly
exhibits double-peaked behavior.  The plasma clumps in this model are
180$^\circ$ apart because the deepest local minima in the effective
potential exist along the line of intersection between the rotational
and magnetic planes (see Equation~22 of \citealt{Townsend2005}).

@@

Microphysically, it is not obvious whether hydrogen alone can produce
the broadband chromatic flux variations seen in CPVs.  The alternative
is that charged dust could provide most of the opacity
\citep{Sanderson2023}.  For TIC\,141146667, Figures~\ref{fig:fulllc}
and~\ref{fig:spec} show that the transits of the inner H$\alpha$ clump
only partially overlap the complex photometric modulation.  This
implies that some of the complex photometric dips must involve
additional opacity sources or spatially distinct structures.

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.46\textwidth]{f5a.pdf}
		}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{f5b.pdf}
		}
	\end{center}
	\vspace{-0.3cm}
  \caption{{\bf TIC\,141146667 age diagnostics.}
  {\bf a,} Dereddened Gaia DR3 color vs.~absolute magnitude for
  TIC\,141146667, USco (8\,Myr), IC\,2602 (40\,Myr), the Pleiades
  (112\,Myr) and stars within 100\,pc. 
  The location of TIC\,141146667 in this diagram suggests an age of
  30-150\,Myr. 
  {\bf b,} An upper limit on photospheric lithium for TIC\,141146667
  (green triangle) yields a lower bound on the star's age of
  $\gtrsim$20\,Myr.  Comparison stars are from the Gaia-ESO survey
  \citep{Jeffries2023}; rich clusters in these data include NGC\,2264
  (4.5\,Myr), $\lambda$~Ori (8.7\,Myr), ${\rm \gamma}^2$\,Vel
  (16.4\,Myr), NGC\,2547 (35.0\,Myr), IC\,2391 (42.0\,Myr),
  and NGC\,2451A (50.0\,Myr).\added{ Measurement uncertainties are
  large below $+$50\,m\AA; negative EW does not correspond to significant
  emission.}}
  \label{fig:age_diagnostics}
\end{figure}

>>>>

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.925\textwidth]{f2.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Photometric evolution of TIC\,141146667 during the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps were caused by stray light from the Earth (BTJD 3356-3358.5) and
  Moon (BTJD 3359.5-3360.5).  Raw two minute data are in gray; black
  time-averages to ten minute sampling.  {\bf b-c,} Folded TESS light
  curve before and after spectroscopy.  Black now phase-averages to
  100 points per 3.93\ hour cycle.  During BTJD 3352-3356, a state
  switch occurred near BTJD 3353, and the dip at $\phi$$\approx$0.8
  \added{became less pronounced}.  While the \added{dip spanning
  $\phi$=0.6-1.1} was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{f3.pdf}
  \vspace{-0.3cm}
  \captionsetup{labelformat=moviefmt,labelsep=colon}
	\caption{\textbf{Emission from plasma clumps orbiting
TIC\,141146667.}
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{{\bf
  available here}},
  shows the spectral evolution over five hours.
  {\bf a,} TESS light curve from 5 February 2024 to 26
  February 2024 folded on the 3.93\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra from 17 February 2024.
  The continuum is set to unity, and the darkest color is at twice the
  continuum to accentuate emission outside the line core ($|v/v_{\rm
  eq}|>1$, for the equatorial velocity $v_{\rm eq}$=130\,\kms).  While
  emission within the line core comes from the star's chromosphere, the
  sinusoidal emission features are most readily described by a warped
  plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the online movie.
  The dotted line shows the time-averaged spectrum, $f_{\langle t
  \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting $f_{\langle t
  \rangle}$.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch mirrors the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.  \added{A separate version with best-fit sinusoids overplotted is also
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel_sinusoids.mp4}{{\bf
  available}}.}}
  \label{fig:spec}
\end{figure*}

@@

In a follow-up study, \citet{Townsend2008} made synthetic light
curves, assuming that the optical depth scaled linearly with plasma
density.  They found that W-shaped eclipses, similar to those seen for
TIC\,141146667, can occur when the spin and magnetic axes are
moderately (15-45$^\circ$) misaligned.  The movies associated with
their work are available
online\footnote{\url{http://user.astro.wisc.edu/\~townsend/static.php?ref=rrm-movies\#Download\_Bundles}
last accessed 28 May 2025.}; for a $\omega/\omega_{\rm
c}$$\approx$0.5 edge-on system like TIC\,141146667, the photometric
eclipses have the correct shape, and the H$\alpha$ emission similarly
exhibits double-peaked behavior.  The plasma clumps in this model are
180$^\circ$ apart because the deepest local minima in the effective
potential exist along the line of intersection between the rotational
and magnetic planes (see Equation~22 of \citealt{Townsend2005}).

@@

Microphysically, it is not obvious whether hydrogen alone can produce
the chromatic broadband flux variations seen in CPVs.  The alternative
is that charged dust could provide most of the opacity
\citep{Sanderson2023}.  For TIC\,141146667, Figures~\ref{fig:fulllc}
and~\ref{fig:spec} show that the transits of the inner H$\alpha$ clump
only partially overlap the complex photometric modulation.  This
implies that some of the complex photometric dips must involve
additional opacity sources or spatially distinct structures.

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.46\textwidth]{f5a.pdf}
		}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{f5b.pdf}
		}
	\end{center}
	\vspace{-0.3cm}
  \caption{{\bf TIC\,141146667 age diagnostics.}
  {\bf a,} Dereddened Gaia DR3 color vs.~absolute magnitude for
  TIC\,141146667, USco (8\,Myr), IC\,2602 (40\,Myr), the Pleiades
  (112\,Myr) and stars within 100\,pc. 
  The location of TIC\,141146667 in this diagram suggests an age of
  30-150\,Myr. 
  {\bf b,} A dearth of photospheric lithium for TIC\,141146667 (yellow
  triangle denotes 2$\sigma$ upper limit) yields a lower bound on the
  star's age of $\gtrsim$20\,Myr.  Comparison stars are from the
  Gaia-ESO survey \citep{Jeffries2023}; rich clusters in these data
  include NGC\,2264 (4.5\,Myr), $\lambda$~Ori (8.7\,Myr), ${\rm
  \gamma}^2$\,Vel (16.4\,Myr), NGC\,2547 (35.0\,Myr), IC\,2391
  (42.0\,Myr), and NGC\,2451A (50.0\,Myr).\added{ Measurement
  uncertainties are large below $+$50\,m\AA; negative EW does not
  correspond to significant emission.}}
  \label{fig:age_diagnostics}
\end{figure}
